Some examples in extended logic counting classes

### 2017-04-07 LICS/syntactic-clo.tex
minor corrections

It was proved in \cite{SalujaST95} that every function in $\E{1}$ admits an FPRAS. Besides, it can be proved that $\E{1} \subseteq \totp$. 
However, 
%we have to discard 
this class 
%since it i
is not closed under sum, and then it is not robust under basic closure properties. 
\begin{proposition}\label{prop-e1-nc}
There exist functions $f, g \in \E{1}$ such that $(f + g) \not\in \E{1}$.
\end{proposition}
To overcome this limitation, one can consider the class $\QE{1}$, which is closed under sum by definition. In fact, the following proposition shows that the same good properties as for $\E{1}$ hold for $\QE{1}$, together with the fact that 
%plus being 
it is closed under sum and multiplication.
\begin{proposition} \label{prop:qe0-fp-qe1-totp-fptras}
$\QE{1} \subseteq \totp$ and every function in $\QE{1}$ has an FPRAS. Moreover, $\QE{1}$ is closed under sum and multiplication.
\end{proposition}
%Moreover, from the following theorem one can conclude that $\QE{1}$ is closed under multiplication.
%\begin{theorem}\label{theo:binary-prod}
%	If $\LL$ is closed under conjunction, then $\eqso(\LL)$ is closed under multiplication.
%\end{theorem}
Hence, it only remains to prove that $\QE{1}$ is closed under subtraction by one. Unfortunately, it is not clear whether this property holds; in fact, we conjecture that it is not the case. Thus, we need to find an extension of $\QE{1}$ that keeps all the previous properties and is closed under subtraction by one. It is important to notice that $\shp$ is believed not to be closed under subtraction by one by some complexity-theoretical assumption.\footnote{A decision problem $L$ is in the randomized complexity class $\cspp$ if there exists a polynomial-time NTM $M$ such that for every $x \in L$ it holds that $\tma_M(x) - \tmr_M(x) = 2$, and for every $x \not\in L$ it holds that $\tma_M(x) = \tmr_M(x)$ \cite{OH93,FFK94}. It is believed that $\np \not\subseteq \cspp$.
%, as there is no evidence that a TM with the property just described exists for the propositional satisfiability problem. 
However, if $\shp$ is closed under subtraction by one, then it holds that $\np \subseteq \cspp$ \cite{OH93}.} So, the following proposition rules out any logic that extends $\Pi_1$ for a possible extension of~$\QE{1}$ with the desired closure property.
\begin{proposition} \label{pi-minusone}
If $\Pi_1 \subseteq \LL \subseteq \fo$ and $\eqso(\LL)$ is closed under subtraction by one, then $\shp$ is closed under subtraction by~one. 
\end{proposition}
Therefore, the desired extension has to be achieved by allowing some local extensions to $\Sigma_1$. More precisely, we define $\logex{1}$ as $\Sigma_1$ but allowing atomic formulae over a signature $\R$ to be of the form either $u = v$ or $X(\bar u)$, where $X$ is a second-order variable, or $\varphi(\bar u)$, where $\varphi(\bar u)$ is a first-order formula over $\R$ (in particular, it does not mentioned any second-order variable). With this extension we obtain a class with the desired properties.
%find the promised class.
\begin{theorem}\label{sigmafo-minusone}
The class $\eqso(\logex{1})$ is closed under sum, multiplication and subtraction by one. Moreover, $\eqso(\logex{1}) \subseteq \totp$ and every function in $\eqso(\logex{1})$ has an FPRAS.
\end{theorem}
The proof that $\eqso(\logex{1})$ is closed under subtraction by one is the most involved of the paper. We think the main technique used in this proof, which is based on considering some witnesses of logarithmic size, is of independent interest.

>>>>

It was proved in \cite{SalujaST95} that every function in $\E{1}$ admits an FPRAS. Besides, it can be proved that $\E{1} \subseteq \totp$. 
However, 
%we have to discard 
this class 
%since it i
is not closed under sum, and then it is not robust under basic closure properties. 
\begin{proposition}\label{prop-e1-nc}
There exist functions $f, g \in \E{1}$ such that $(f + g) \not\in \E{1}$.
\end{proposition}
To overcome this limitation, one can consider the class $\QE{1}$, which is closed under sum by definition. In fact, the following proposition shows that the same good properties as for $\E{1}$ hold for $\QE{1}$, together with the fact that 
%plus being 
it is closed under sum and multiplication.
\begin{proposition} \label{prop:qe0-fp-qe1-totp-fptras}
$\QE{1} \subseteq \totp$ and every function in $\QE{1}$ has an FPRAS. Moreover, $\QE{1}$ is closed under sum and multiplication.
\end{proposition}
%Moreover, from the following theorem one can conclude that $\QE{1}$ is closed under multiplication.
%\begin{theorem}\label{theo:binary-prod}
%	If $\LL$ is closed under conjunction, then $\eqso(\LL)$ is closed under multiplication.
%\end{theorem}
Hence, it only remains to prove that $\QE{1}$ is closed under subtraction by one. Unfortunately, it is not clear whether this property holds; in fact, we conjecture that it is not the case. Thus, we need to find an extension of $\QE{1}$ that keeps all the previous properties and is closed under subtraction by one. It is important to notice that $\shp$ is believed not to be closed under subtraction by one by some complexity-theoretical assumption.\footnote{A decision problem $L$ is in the randomized complexity class $\cspp$ if there exists a polynomial-time NTM $M$ such that for every $x \in L$ it holds that $\tma_M(x) - \tmr_M(x) = 2$, and for every $x \not\in L$ it holds that $\tma_M(x) = \tmr_M(x)$ \cite{OH93,FFK94}. It is believed that $\np \not\subseteq \cspp$.
%, as there is no evidence that a TM with the property just described exists for the propositional satisfiability problem. 
However, if $\shp$ is closed under subtraction by one, then it holds that $\np \subseteq \cspp$ \cite{OH93}.} So, the following proposition rules out any logic that extends $\Pi_1$ for a possible extension of~$\QE{1}$ with the desired closure property.
\begin{proposition} \label{pi-minusone}
If $\Pi_1 \subseteq \LL \subseteq \fo$ and $\eqso(\LL)$ is closed under subtraction by one, then $\shp$ is closed under subtraction by~one. 
\end{proposition}
Therefore, the desired extension has to be achieved by allowing some local extensions to $\Sigma_1$. More precisely, we define $\logex{1}$ as $\Sigma_1$ but allowing atomic formulae over a signature $\R$ to be of the form either $u = v$ or $X(\bar u)$, where $X$ is a second-order variable, or $\varphi(\bar u)$, where $\varphi(\bar u)$ is a first-order formula over $\R$ (in particular, it does not mention any second-order variable). With this extension we obtain a class with the desired properties.
%find the promised class.
\begin{theorem}\label{sigmafo-minusone}
The class $\eqso(\logex{1})$ is closed under sum, multiplication and subtraction by one. Moreover, $\eqso(\logex{1}) \subseteq \totp$ and every function in $\eqso(\logex{1})$ has an FPRAS.
\end{theorem}
The proof that $\eqso(\logex{1})$ is closed under subtraction by one is the most involved of the paper. We think the main technique used in this proof, which is based on considering some witnesses of logarithmic size, is of independent interest.

### 2017-04-07 LICS/syntactic.tex
minor corrections

Counting problems with easy decision versions play a fundamental role in the search of efficient approximations algorithms for functions in $\shp$. A fully-polynomial randomized approximation scheme (FPRAS) for a function $f : \Sigma^* \to \bbN$ is a randomized algorithm ${\cal A} : \Sigma^* \times (0,1) \to \bbN$ such that: (1) for every string $x \in \Sigma^*$ and real value $\varepsilon \in (0,1)$, the probability that $|f(x) - {\cal A}(x,\varepsilon)| \leq \varepsilon \cdot f(x)$ is at least $\frac{3}{4}$, and (2) the running time of ${\cal A}$ is polynomial in the size of $x$ and $1/\varepsilon$ \cite{KL83}. Notably, there exist $\shp$-complete functions that can be efficiently approximated as they admit FPRAS; for instance, there exist FPRAS for $\cdnf$~\cite{KL83} and $\cpm$~\cite{JSV04}. A key observation here is that if a function $f$ admits an FPRAS, then $L_f$ is in the randomized complexity class $\bpp$~\cite{G77}. Hence, under the widely believed assumption that  $\np \not\subseteq \bpp$, we cannot hope for an FPRAS for a function in $\shp$ whose decision counterpart is $\np$-complete, and we have to concentrate on the class of counting problems with easy decision versions.

The importance of the class of counting problems with easy decision counterparts has motivated the search of robust classes of functions in $\shp$ with this property \cite{PagourtzisZ06}. But the key question here is what should be considered a {\em robust} class. 
A first desirable condition has to do with the closure properties satisfied by the class, which is a common theme when studying function complexity classes \cite{OH93,FH08}. As in the cases of $\ptime$ and $\np$ that are closed under intersection and union, we expect our class to be closed under multiplication and sum. For a more elaborated closure property, assume that $\textit{sat\_one}$ is a function that returns one plus the number of satisfying assignments of a propositional formula. Clearly $\textit{sat\_one}$ is a $\shp$-complete function whose decision counterpart $L_{\textit{sat\_one}}$ is trivial. But should $\textit{sat\_one}$ be part of a robust class of counting function with easy decision versions? The key insight here is that if a function in $\shp$ has an easy decision counterpart $L$, then as $L \in \np$ we expect  to have a polynomial-time algorithm that verifies whether $x \in L$ by constructing witnesses for~$x$. 
%As an example of this considered all the counting functions mentioned in the previous paragraphs and, in particular, the decision counterpart of $\cdnf$ that can be solved by explicitly constructing in polynomial-time a satisfying assignment for an input propositional formula. 
Moreover, if such an algorithm for constructing witnesses exists, then we also expect to be able to manipulate such witnesses and in some cases to remove them. In other words, we expect a robust class $\CC$ of counting functions with easy decision versions to be closed under subtraction by one, that is, if $g \in \CC$, then the function $g \dotdiv 1$ should also be in $\CC$, where $(g \dotdiv 1)(x)$ is defined as $g(x) - 1$ if $g(x) \geq 1$, and as~$0$ otherwise. Notice that, unless $\ptime = \np$, no such class can contain the function $\textit{sat\_one}$ because $\textit{sat\_one} \dotdiv 1$ counts the number of satisfying assignments of a propositional formula.

A second desirable condition of robustness is the existence of natural complete problems \cite{P94}. Special attention has to be paid here to the notion of reduction used for completeness. Notice that under the notion of Cook reduction, originally used in \cite{Valiant79}, the problems $\cdnf$ and $\csat$ are $\shp$-complete. However, $\cdnf$ has an easy decision counterpart and admits an FPRAS, while $\csat$ does not satisfy these conditions unless $\ptime = \np$. Hence a more strict notion of reduction has to be considered; in particular, the notion of parsimonious reduction (to be defined later) satisfy that if a function $f$ is parsimoniously reducible to a function $g$, then $L_g \in \ptime$ implies that $L_f \in \ptime$ and the existence of an FPRAS for $g$ implies the existence of a FPRAS for $f$.

>>>>

Counting problems with easy decision versions play a fundamental role in the search of efficient approximation algorithms for functions in $\shp$. A fully-polynomial randomized approximation scheme (FPRAS) for a function $f : \Sigma^* \to \bbN$ is a randomized algorithm ${\cal A} : \Sigma^* \times (0,1) \to \bbN$ such that: (1) for every string $x \in \Sigma^*$ and real value $\varepsilon \in (0,1)$, the probability that $|f(x) - {\cal A}(x,\varepsilon)| \leq \varepsilon \cdot f(x)$ is at least $\frac{3}{4}$, and (2) the running time of ${\cal A}$ is polynomial in the size of $x$ and $1/\varepsilon$ \cite{KL83}. Notably, there exist $\shp$-complete functions that can be efficiently approximated as they admit FPRAS; for instance, there exist FPRAS for $\cdnf$~\cite{KL83} and $\cpm$~\cite{JSV04}. A key observation here is that if a function $f$ admits an FPRAS, then $L_f$ is in the randomized complexity class $\bpp$~\cite{G77}. Hence, under the widely believed assumption that  $\np \not\subseteq \bpp$, we cannot hope for an FPRAS for a function in $\shp$ whose decision counterpart is $\np$-complete, and we have to concentrate on the class of counting problems with easy decision versions.

The importance of the class of counting problems with easy decision counterparts has motivated the search of robust classes of functions in $\shp$ with this property \cite{PagourtzisZ06}. But the key question here is what should be considered a {\em robust} class. 
A first desirable condition has to do with the closure properties satisfied by the class, which is a common theme when studying function complexity classes \cite{OH93,FH08}. As in the cases of $\ptime$ and $\np$ that are closed under intersection and union, we expect our class to be closed under multiplication and sum. For a more elaborated closure property, assume that $\textit{sat\_one}$ is a function that returns one plus the number of satisfying assignments of a propositional formula. Clearly $\textit{sat\_one}$ is a $\shp$-complete function whose decision counterpart $L_{\textit{sat\_one}}$ is trivial. But should $\textit{sat\_one}$ be part of a robust class of counting functions with easy decision versions? The key insight here is that if a function in $\shp$ has an easy decision counterpart $L$, then as $L \in \np$ we expect  to have a polynomial-time algorithm that verifies whether $x \in L$ by constructing witnesses for~$x$. 
%As an example of this considered all the counting functions mentioned in the previous paragraphs and, in particular, the decision counterpart of $\cdnf$ that can be solved by explicitly constructing in polynomial-time a satisfying assignment for an input propositional formula. 
Moreover, if such an algorithm for constructing witnesses exists, then we also expect to be able to manipulate such witnesses and in some cases to remove them. In other words, we expect a robust class $\CC$ of counting functions with easy decision versions to be closed under subtraction by one, that is, if $g \in \CC$, then the function $g \dotdiv 1$ should also be in $\CC$, where $(g \dotdiv 1)(x)$ is defined as $g(x) - 1$ if $g(x) \geq 1$, and as~$0$ otherwise. Notice that, unless $\ptime = \np$, no such class can contain the function $\textit{sat\_one}$ because $\textit{sat\_one} \dotdiv 1$ counts the number of satisfying assignments of a propositional formula.

A second desirable condition of robustness is the existence of natural complete problems \cite{P94}. Special attention has to be paid here to the notion of reduction used for completeness. Notice that under the notion of Cook reduction, originally used in \cite{Valiant79}, the problems $\cdnf$ and $\csat$ are $\shp$-complete. However, $\cdnf$ has an easy decision counterpart and admits an FPRAS, while $\csat$ does not satisfy these conditions unless $\ptime = \np$. Hence a more strict notion of reduction has to be considered; in particular, the notion of parsimonious reduction (to be defined later) satisfies that if a function $f$ is parsimoniously reducible to a function $g$, then $L_g \in \ptime$ implies that $L_f \in \ptime$ and the existence of an FPRAS for $g$ implies the existence of a FPRAS for $f$.
